Conversational expectations account for apparent limits on theory of mind use

### 2016-01-29 writing/cogsci-submission/cogsci_keysar.tex
rearranged conclusion a bit

Pragmatic language understanding requires sophisticated social reasoning. To interpret an utterance, a listener must consider how a speaker is likely to behave in context.
The rational use of theory of mind for a listener thus depends on her expectations about the speaker: If, in a particular context, she expects a speaker to provide sufficiently informative utterances, she may be justified in neglecting his epistemic state.

%consider the speaker's communicative goals and context. 
In both experiments, we found evidence that listeners neglect the speaker's epistemic state, as Keysar and colleagues claim.
More specifically, we found that when the speaker's utterance fit the hidden distractor better than the target, matchers were more likely to make errors. This suggests that reference disambiguation was driven primarily by \emph{a priori} label fitness. 
%
However, this does not necessarily imply limitations on theory of mind---if speakers would naturally provide expressions which apply better to the target, then this behavior would be appropriate.
% If listeners arrive at this interpretation strategy through pragmatic reasoning, then neglecting the speaker's epistemic state would be rational and also the result of theory of mind use. 
In Exp.~2, we found that speakers did naturally produce more precise, informative utterances than required and these unconstrained utterances fit the target significantly better than they fit the hidden distractor. Thus, the patterns of errors observed in Exp.~1 can be explained as an uncooperative confederate---the experiment set up certain pragmatic expectations through the task context, then deliberately flouted them in critical trials. % Removing constraints on the speaker produced a dramatic reduction in overall matcher errors and systematic variation in error rates across items. We found that unscripted speakers systematically tended to produce more precise, costly referring expressions than their scripted counterparts -- for example, ``the clear audiocassette'' when there was only one ``tape'' in the speaker's display. These precise labels were less likely to fit the hidden distractor, and therefore less likely to lead to a matcher error.

@@

\ndg{there's an abrupt jump here. play with ordering?}

The Gricean maxim of quality dictates that cooperative speakers should make their utterance as informative as is required for current purposes, but no more so. If the referring expressions used in Exp.~1 uniquely pick out the target from the speaker's perspective, then why would unscripted speakers in Exp.~2 strive to be more informative than strictly necessary and why would listeners come to expect such ``overinformativity''? Studies of overinformativity in referring expressions have uncovered a number of factors, such as the tendency to mention perceptually salient features to speed up the identification process  \cite{KoolenGattGoudbeekKrahmer11_Overspecification}. In the current task, it seems particularly relevant that both participants knew about the presence of hidden objects, increasing uncertainty about possible distractors and placing additional pressure on the speaker to overspecify.

It is interesting that the item-wise variability we observed in both experiments is at least partially due to the availability and effectiveness of more specific expressions. For example, no speaker in Exp.~2 produced ``the bottom block,'' which was used as the critical instruction in Exp.~1. Instead, they said ``the bluish block with a B'' or ``block with the blue writing,'' which relied on less confusable perceptual features and thereby decreased error rates. The items on which errors were most likely in Exp.~2 were difficult precisely because the overinformative utterance was \emph{still} ambiguous: even if the speaker uses ``hairbrush'' instead of the basic-level ``brush,'' for example, the distractor remains a more typical exemplar of the subcategory.

\ndg{concluding paragraph. Overall, we both replicated the finding that listeners neglect the speakers epistemic state, and explained why this may be so (and why errors are uneven over items) ...
}

>>>>

Pragmatic language understanding requires sophisticated social reasoning. To interpret an utterance, a listener must consider how a speaker is likely to behave in context.
The rational use of theory of mind for a listener thus depends on her expectations about the speaker: If, in a particular context, she expects a speaker to provide sufficiently informative utterances, she may be justified in neglecting his epistemic state when resolving reference.

%consider the speaker's communicative goals and context. 
In our replication (Expt.~1), we found evidence that listeners neglect the speaker's epistemic state, as Keysar and colleagues claim.
We also found that when the speaker's utterance fit the hidden distractor better than the target, matchers were more likely to make errors. 
Indeed, we found that the extremely heterogeneous pattern of errors across items was well explained by the relative fit of the utterance to target and distractor.
This suggests that reference disambiguation was driven primarily by \emph{a priori} label fitness. 
%
However, this does not necessarily imply limitations on theory of mind---if speakers could be expected to naturally provide expressions which apply better to the target, then this behavior would be appropriate.
% If listeners arrive at this interpretation strategy through pragmatic reasoning, then neglecting the speaker's epistemic state would be rational and also the result of theory of mind use. 
In Expt.~2, we found that speakers did naturally produce more precise, informative utterances than required and these unconstrained utterances fit the target significantly better than they fit the hidden distractor. 
For example, no speaker in Expt.~2 produced ``the bottom block,'' which was used as a critical instruction in Exp.~1. Instead, they said ``the bluish block with a B'' or ``block with the blue writing,'' which relied on less confusable perceptual features and thereby decreased error rates. 
Thus, the errors observed in Expt.~1 can be explained as a result of an uncooperative confederate (speaker)---the experiment set up certain pragmatic expectations through the task context, then deliberately flouted them in critical trials. % Removing constraints on the speaker produced a dramatic reduction in overall matcher errors and systematic variation in error rates across items. We found that unscripted speakers systematically tended to produce more precise, costly referring expressions than their scripted counterparts -- for example, ``the clear audiocassette'' when there was only one ``tape'' in the speaker's display. These precise labels were less likely to fit the hidden distractor, and therefore less likely to lead to a matcher error.

The Gricean maxim of quality dictates that cooperative speakers should make their utterance as informative as is required for current purposes, but no more so. If the referring expressions used in Expt.~1 uniquely pick out the target from the speaker's perspective, then why would unscripted speakers in Expt.~2 be more informative than this? 
%strictly necessary and why would listeners come to expect such ``overinformativity''? 
One possibility is that awareness of the complex mismatch in epistemic state with the listener leads the speaker to provide extra information in an attempt to avoid miscommunication.
Another possibility is that more general dynamics of language are in play.
Studies of over-informativity in referring expressions have uncovered a general tendency of speakers to provide redundant information.
This tendency can depend on a number of situational factors, such as the tendency to mention perceptually salient features to speed up the identification process  \cite{KoolenGattGoudbeekKrahmer11_Overspecification}. 
%In the current task, it seems particularly relevant that both participants knew about the presence of hidden objects, increasing uncertainty about possible distractors and placing additional pressure on the speaker to overspecify. 
The exact origin is a subject that must be followed up by future research, but it is clear that the tendency of speakers to produce highly informative referring expressions is useful to, and relied on by, listeners.
Indeed, the remaining errors in Expt.~2 are at least partially due to the lack of availability of more specific expressions. 
The items on which errors were most likely in Expt.~2 were difficult precisely because the natural utterance was \emph{still} ambiguous: even if the speaker uses ``hairbrush'' instead of the basic-level ``brush,'' for example, the distractor remains a more typical exemplar of the subcategory.

@@

\ndg{concluding paragraph? Overall, we both replicated the finding that listeners neglect the speakers epistemic state, and explained why this may be so (and why errors are uneven over items) ...
Or maybe we don't need anything else.
}
